### 2021-09-26 documents/sections/verbs.tex
proofread 4.6

\subsection{\rc{(ət-)jəpɨ} \qu{to come}}
\label{sec:come}
\definecolor{come1}{RGB}{161.0,201.0,244.0}
\definecolor{come2}{RGB}{255.0,180.0,130.0}
\definecolor{come3}{RGB}{141.0,229.0,161.0}
\definecolor{come4}{RGB}{255.0,159.0,155.0}
\begin{table}
	\centering
	\caption{Reflexes of \qu{to come}}
    \label{tab:come}
	\begin{tabular}{@{}lll@{}}
	\mytoprule
	Language & Form & Source\\
	\mymidrule
\kaxui & \colorbox{come3}{\obj{ehɨ}} & \perscomm{Spike Gildea}\\
\kaxui & \colorbox{come1}{\obj{johɨ}} & \perscomm{Spike Gildea}\\
\kaxui & \colorbox{come2}{\obj{oohɨ}} & \perscomm{Spike Gildea}\\
\arara & \colorbox{come3}{\obj{ebɨ}} & \textcite[150]{alves2017arara}\\
\arara & \colorbox{come4}{\obj{odebɨ}} & \textcite[150]{alves2017arara}\\
\ikpeng & \colorbox{come4}{\obj{arep}} & \textcite[265]{ikpengpacheco2001}\\
\bakairi & \colorbox{come4}{\obj{əewɨ}} & \textcite[4]{meira2003bakairi}\\
\trio & \colorbox{come4}{\obj{əepɨ}} & \textcite[294]{triomeira1999}\\
\trio & \colorbox{come3}{\obj{epɨ}} & \textcite[294]{triomeira1999}\\
\akuriyo & \colorbox{come3}{\obj{epɨ}} & \textcite[168]{meira1998proto}\\
\carijo & \colorbox{come3}{\obj{ehɨ}} & \textcite[178]{robayo2000avance}\\
\apalai & \colorbox{come4}{\obj{oepɨ}} & \textcite[37]{koehn1986apalai}\\
\kalina & \colorbox{come2}{\obj{opɨ}} & \textcite[429]{courtz2008carib}\\
\maqui & \colorbox{come3}{\obj{ehə}} & \textcite[438]{maquiritaricaceres2011}\\
\akawaio & \colorbox{come4}{\obj{əsipɨ}} & \textcite[160]{stegeman2014akawaio}\\
\akawaio & \colorbox{come1}{\obj{jepɨ}} & \textcite[125]{akawaiocaesar2003}\\
\ingariko & \colorbox{come1}{\obj{jepə}} & \textcite[415]{cruz2005fonologia}\\
\ingariko & \colorbox{come1}{\obj{jə}} & \textcite[299]{cruz2005fonologia}\\
\patamona & \colorbox{come1}{\obj{jepɨ}} & \perscomm{Spike Gildea}\\
\patamona & \colorbox{come1}{\obj{jəpɨ}} & \perscomm{Spike Gildea}\\
\pemon & \colorbox{come1}{\obj{jepɨ}} & \textcite[102]{alvarez2000construcciones}\\
\macushi & \colorbox{come1}{\obj{ipɨ}} & \textcite[32]{macushiabbott1991}\\
\panare & \colorbox{come2}{\obj{əpɨ}} & \textcite[65]{panarepayne2013}\\
\yawarana & \colorbox{come2}{\obj{əpɨ}} & \textcite[68]{mendez1959yawarana}\\
\mapoyo & \colorbox{come3}{\obj{epɨ}} & \textcite[74]{muller1975mapoyo}\\
\uxc & \colorbox{come2}{\obj{ee}} & \textcite[182]{meira2005southern}\\
\mybottomrule
	\end{tabular}
\end{table}
This verb is reconstructed as monomorphemic \rc{ətepɨ} by \textcite[30]{gildea2007greenberg}, but an inspection of all the attested reflexes \pcref{tab:come} suggests a more complex story.
In particular, the Pemongan languages and one of its allomorphs in \kaxui suggest a root \colorbox{come1}{\rc{jəpɨ}}.
However, the one form in \kaxui only occurs in the Progressive \exref{kaxprog.kax-82}, and \obj{j(-)} may be a reflex of the \settwo third person marker \rc{i-} and the \gl{s_a_} class marker \rc{w-}, as suggested by the occurrence of the latter with other person values \exref{kaxprog.kax-54}.
On the other hand, while C-initial verbs do show a clear reflex of third person \rc{i-} \exref{kaxprog.kax-77}, regular V-initial verbs do not show \obj{j-}, but Ø \exref{kaxprog.kax-83}.
Thus, it seems more likely that the \obj{j} is indeed part of the verb, rather than an outcome of \rc{i-w-}.

@@

I have so far ignored the forms which form the basis for \posscite{gildea2007greenberg} \rc{ətepɨ}; these are compatible with my reconstruction \rc{jəpɨ}, with the addition of the detransitivizer \rc{ət(e)-}: \colorbox{come4}{\rc{ətjəpɨ}}.
In fact, the \obj{i} in the \akawaio form \obj{əsipɨ} is better explained by a sequence \rc{jə}, which has the same outcome in the \macushi reflex of \rc{jəpɨ}, \obj{ipɨ}.
While the semantics of combining a detransitivizer with an intransitive verb do not really make sense, some historical \gl{s_p_} verbs are attested as adding the detransitivizer to become \gl{s_a_} verbs.
For example, \gl{s_p_} \rc{wənɨkɨ} \qu{to sleep} becomes \trio \obj{əənɨkɨ} \parencite[252]{triomeira1999} and \kalina \obj{əʔnɨkɨ} \parencite[429]{courtz2008carib}, both \gl{s_a_}.
\waiwai \qu{go to sleep} can be \obj{wɨnɨk} \parencite[30]{waiwaihawkins1998} or \obj{et-wɨnɨk} \parencite[204]{hawkins1953waiwai}.
The parallels to \qu{to sleep} end here, since bare \rc{jəpɨ} \qu{to come} apparently already was an \gl{s_a_} verb, as evidenced by its status in \kaxui, \kalina, \panare \exref{pan-128}, \arara, \trio, and \akuriyo.

@@

While these sound changes and the addition of \rc{ət(e)-} do account for the majority of the forms in \cref{tab:come},\footnote{The only exception is the \apalai form \obj{oepɨ}, where the detransitivizer would have the reflex \obj{os-} \parencite[506]{meira2010origin}. While \obj{oepɨ} would be a regular outcome of \rc{ə-jəpɨ}, the \envr{}{C} allomorph of the detransitivizer is \obj{e-}. The form may be due to borrowing from \trio, which has lost intervocalic \rc{t} to create \obj{əepɨ}. Alternatively, \apalai \obj{oepɨ} could be from \wayana, which has lost its reflex of \rc{ətjəpɨ}, but where regular sound changes would also have resulted in the loss of intervocalic \rc{t} \parencite[63]{wayanatavares2005}.}
the distribution within the family is rather chaotic.
Not only do very closely related languages show different forms, like \yawarana and \mapoyo, but distinct forms can even be found within the same language, usually conditioned by different prefixes.
This was discussed in \cref{sec:pekodian} for \arara, which has reflexes of \rc{ətjəpɨ} and \rc{jəpɨ} within the same paradigm.
A similar situation is found in \trio, where the \setone paradigm shows a reflex of \rc{ətjəpɨ} for first, but of \rc{epɨ} (< \rc{jəpɨ}) for the other persons \exref{tricome}.\footnote{While the \gl{1+2} form is a regular outcome of \rc{kɨt-epɨ}, the second person form is mysterious.}

@@

\subsection{\rc{ɨpɨtə} \qu{to go down}}
\label{sec:godown}
Reflexes of this verb were not affected by the extensions of \rc{t-} in \PTir \pcref{sec:taranoan} and \obj{k-} in \akuriyo \pcref{sec:akuriyo}.
Rather, an irregular first person form \rc{p-ɨptə} can be reconstructed for \PTir.
While it seems to have resisted the extension of \rc{k-} in \PPek, \bakairi subsequently introduced it \pcref{sec:pekodian}.
It was also affected by the extension of \obj{j(i)-} in \carijo \exref{car-32}.

@@

A comparative view \pcref{tab:descend_cog} shows that while a form \rc{ɨpɨtə} can be reconstructed to \PC, the class of this verb is unclear at first sight.
The reflexes are fairly evenly split between \gl{s_a_} and \gl{s_p_} -- in those languages that preserve the split-\gl{s} system.
In one language, \wayana, the verb shows traits of both classes, leading me to consider it a \dbqu{mixed} verb.
It takes the first and second person \gl{s_p_} markers \obj{j-} and \obj{əw-} \parencite[200]{wayanatavares2005}, but the \gl{1+2}\gl{s_a_} marker \obj{kut-} \parencite[206]{wayanatavares2005}.
It also shows the \gl{s_a_} class marker \obj{w-} in nominalizations \exref{waygodown.way-73}, but behaves like an \gl{s_p_} verb in taking a second person prefix in imperatives \exref{waygodown.way-71}.

\pex<waygodown>\wayana \parencite[][200]{wayanatavares2005}
\a<way-73>
\begingl
\glpreamble ïwïptëë//
\gla ɨ-w-ɨptə-rɨ//
\glb \gl{1}-\gl{s_a_}-go.down-\gl{nmlz}//
\glft \qu{my going down}//
\endgl
\a<way-71>
\begingl
\gla əw-ɨptə-k//
\glb \gl{2}-go.down-\gl{imp}//
\glft \qu{Go down!}//
\endgl
\xe
%
Its causativized form is \obj{ɨptəka} \parencite[255]{wayanatavares2005}, containing a reflex of the transitivizer \rc{-ka}, which was restricted to \gl{s_p_} verbs \parencite{gildea2019overview}.

Comparative evidence suggests that \rc{ɨpɨtə} \qu{to go down} was originally an \gl{s_p_} verb which switched its class to \gl{s_a_} in some languages, incompletely so in \wayana.
The \arara causativized form is \obj{eniptoŋ} \parencite[66]{alves2017arara}, and \kalina has a cognate form \obj{enɨʔto} \parencite[263]{courtz2008carib}; \obj{onɨʔto} \qu{to go down} in \cref{tab:descend_cog} is a detransitivized form thereof, lit.\ \qu{to get oneself down}.
Both causativized forms contain a reflex of the (rare) transitivizer \rc{en-}, which was usually found with \gl{s_p_} verbs \parencite{gildea2019overview}.
Besides the irregular first person \obj{p-}, \trio \obj{ɨhtə} shows other irregularities, in particular in its causativized forms \parencite[263]{triomeira1999}.
Thus, it seems that this verb was originally \gl{s_p_}, but then switched its class in four and a half languages of the family, for so far unknown reasons.

@@

>>>>

\subsection{\rc{(ət-)jəpɨ} \qu{to come}}
\label{sec:come}
\definecolor{come1}{RGB}{161.0,201.0,244.0}
\definecolor{come2}{RGB}{255.0,180.0,130.0}
\definecolor{come3}{RGB}{141.0,229.0,161.0}
\definecolor{come4}{RGB}{255.0,159.0,155.0}
\begin{table}
	\centering
	\caption{Reflexes of \qu{to come}}
    \label{tab:come}
	\begin{tabular}{@{}lll@{}}
	\mytoprule
	Language & Form & Source\\
	\mymidrule
\kaxui & \colorbox{come3}{\obj{ehɨ}} & \perscomm{Spike Gildea}\\
\kaxui & \colorbox{come1}{\obj{johɨ}} & \perscomm{Spike Gildea}\\
\kaxui & \colorbox{come2}{\obj{oohɨ}} & \perscomm{Spike Gildea}\\
\arara & \colorbox{come3}{\obj{ebɨ}} & \textcite[150]{alves2017arara}\\
\arara & \colorbox{come4}{\obj{odebɨ}} & \textcite[150]{alves2017arara}\\
\ikpeng & \colorbox{come4}{\obj{arep}} & \textcite[265]{ikpengpacheco2001}\\
\bakairi & \colorbox{come4}{\obj{əewɨ}} & \textcite[4]{meira2003bakairi}\\
\trio & \colorbox{come4}{\obj{əepɨ}} & \textcite[294]{triomeira1999}\\
\trio & \colorbox{come3}{\obj{epɨ}} & \textcite[294]{triomeira1999}\\
\akuriyo & \colorbox{come3}{\obj{epɨ}} & \textcite[168]{meira1998proto}\\
\carijo & \colorbox{come3}{\obj{ehɨ}} & \textcite[178]{robayo2000avance}\\
\apalai & \colorbox{come4}{\obj{oepɨ}} & \textcite[37]{koehn1986apalai}\\
\kalina & \colorbox{come2}{\obj{opɨ}} & \textcite[429]{courtz2008carib}\\
\maqui & \colorbox{come3}{\obj{ehə}} & \textcite[438]{maquiritaricaceres2011}\\
\akawaio & \colorbox{come4}{\obj{əsipɨ}} & \textcite[160]{stegeman2014akawaio}\\
\akawaio & \colorbox{come1}{\obj{jepɨ}} & \textcite[125]{akawaiocaesar2003}\\
\ingariko & \colorbox{come1}{\obj{jepə}} & \textcite[415]{cruz2005fonologia}\\
\ingariko & \colorbox{come1}{\obj{jə}} & \textcite[299]{cruz2005fonologia}\\
\patamona & \colorbox{come1}{\obj{jepɨ}} & \perscomm{Spike Gildea}\\
\patamona & \colorbox{come1}{\obj{jəpɨ}} & \perscomm{Spike Gildea}\\
\pemon & \colorbox{come1}{\obj{jepɨ}} & \textcite[102]{alvarez2000construcciones}\\
\macushi & \colorbox{come1}{\obj{ipɨ}} & \textcite[32]{macushiabbott1991}\\
\panare & \colorbox{come2}{\obj{əpɨ}} & \textcite[65]{panarepayne2013}\\
\yawarana & \colorbox{come2}{\obj{əpɨ}} & \textcite[68]{mendez1959yawarana}\\
\mapoyo & \colorbox{come3}{\obj{epɨ}} & \textcite[74]{muller1975mapoyo}\\
\uxc & \colorbox{come2}{\obj{ee}} & \textcite[182]{meira2005southern}\\
\mybottomrule
	\end{tabular}
\end{table}
This verb is reconstructed as monomorphemic \rc{ətepɨ} by \textcite[30]{gildea2007greenberg}, but an inspection of all the attested reflexes \pcref{tab:come} suggests a more complex story.
In particular, the Pemongan languages and one of its allomorphs in \kaxui suggest a root \colorbox{come1}{\rc{jəpɨ}}.
However, the one form in \kaxui only occurs in the Progressive \exref{kaxprog.kax-82}, and \obj{j(-)} may be a reflex the \settwo third person marker \rc{i-} combined with  the \gl{s_a_} class marker \rc{w-}, as suggested by the occurrence of the latter with other person values \exref{kaxprog.kax-54}.
On the other hand, while C-initial verbs do show a clear reflex of third person \rc{i-} \exref{kaxprog.kax-77}, regular V-initial verbs do not show \obj{j-}, but Ø \exref{kaxprog.kax-83}.
Thus, it seems more likely that the \obj{j} is indeed part of the verb, rather than an outcome of \rc{i-w-}.

@@

we have so far ignored the forms which form the basis for \posscite{gildea2007greenberg} \rc{ətepɨ}; these are compatible with our reconstructed root \rc{jəpɨ}, with the addition of the detransitivizer \rc{ət(e)-}: \colorbox{come4}{\rc{ətjəpɨ}}.
In fact, the \obj{i} in the \akawaio form \obj{əsipɨ} is better explained by a sequence \rc{jə}, which has the same outcome \obj{i} in the \macushi reflex of \rc{jəpɨ}, \obj{ipɨ}.
While the semantics of combining a detransitivizer with an intransitive verb do not really make sense, some historical \gl{s_p_} verbs are attested as adding the detransitivizer to become \gl{s_a_} verbs.
For example, \gl{s_p_} \rc{wənɨkɨ} \qu{to sleep} becomes \trio \obj{əənɨkɨ} \parencite[252]{triomeira1999} and \kalina \obj{əʔnɨkɨ} \parencite[429]{courtz2008carib}, both \gl{s_a_}.
\waiwai \qu{go to sleep} can be \obj{wɨnɨk} \parencite[30]{waiwaihawkins1998} or \obj{et-wɨnɨk} \parencite[204]{hawkins1953waiwai}.
The parallels to \qu{to sleep} end here, since bare \rc{jəpɨ} \qu{to come} apparently already was an \gl{s_a_} verb, as evidenced by its status in \kaxui, \kalina, \panare \exref{pan-128}, \arara, \trio, and \akuriyo.

@@

While these sound changes and the addition of \rc{ət(e)-} do account for the majority of the forms in \cref{tab:come},\footnote{The only exception is the \apalai form \obj{oepɨ}, where the detransitivizer would have the reflex \obj{os-} \parencite[506]{meira2010origin}. While \obj{oepɨ} would be a regular outcome of \rc{ə-jəpɨ}, the \envr{}{C} allomorph of the detransitivizer is \obj{e-}. The form may be due to borrowing from \trio, which has lost intervocalic \rc{t} to create \obj{əepɨ}. Alternatively, \apalai \obj{oepɨ} could be a fossilized loan \wayana, which has lost its reflex of \rc{ətjəpɨ}, but where regular sound changes would also have resulted in the loss of intervocalic \rc{t} \parencite[63]{wayanatavares2005}.}
the distribution within the family is rather chaotic.
Not only do very closely related languages show different forms, like \yawarana and \mapoyo, but distinct forms can even be found within the same language, usually conditioned by different prefixes.
This was discussed in \cref{sec:pekodian} for \arara, which has reflexes of \rc{ətjəpɨ} and \rc{jəpɨ} within the same paradigm.
A similar situation is found in \trio, where the \setone paradigm shows a reflex of \rc{ətjəpɨ} for first, but of \rc{epɨ} (< \rc{jəpɨ}) for the other persons \exref{tricome}.\footnote{While the \gl{1+2} form is a regular outcome of \rc{kɨt-epɨ}, the second person form is mysterious.}

@@

\subsection{\rc{ɨpɨtə} \qu{to go down}}
\label{sec:godown}
Reflexes of this verb were not affected by the extensions of \rc{t-} in \PTir \pcref{sec:taranoan} and \obj{k-} in \akuriyo \pcref{sec:akuriyo}.
Rather, an irregular first person form \rc{p-ɨptə} can be reconstructed for \PTir.
While it originally resisted the extension of \rc{k-} in \PPek, \bakairi subsequently introduced it \pcref{sec:pekodian}.
It was also affected by the extension of \obj{j-} in \carijo \exref{car-32}.

@@

@@

A comparative view \pcref{tab:descend_cog} shows that while a form \rc{ɨpɨtə} can be reconstructed to \PC, the class of this verb is unclear at first sight.
The reflexes are fairly evenly split between \gl{s_a_} and \gl{s_p_} -- in those languages that preserve the split-\gl{s} system.
In one language, \wayana, the verb shows traits of both classes, leading us to consider it a \dbqu{mixed} verb.
It takes the first and second person \gl{s_p_} markers \obj{j-} and \obj{əw-} \parencite[200]{wayanatavares2005}, but the \gl{1+2}\gl{s_a_} marker \obj{kut-} \parencite[206]{wayanatavares2005}.
It also shows the \gl{s_a_} class marker \obj{w-} in nominalizations \exref{waygodown.way-73}, but behaves like an \gl{s_p_} verb in taking a second person prefix in imperatives \exref{waygodown.way-71}.

\pex<waygodown>\wayana \parencite[][200]{wayanatavares2005}
\a<way-73>
\begingl
\glpreamble ïwïptëë//
\gla ɨ-w-ɨptə-rɨ//
\glb \gl{1}-\gl{s_a_}-go.down-\gl{nmlz}//
\glft \qu{my going down}//
\endgl
\a<way-71>
\begingl
\gla əw-ɨptə-k//
\glb \gl{2}-go.down-\gl{imp}//
\glft \qu{Go down!}//
\endgl
\xe
%
Its causativized form is \obj{ɨptəka} \parencite[255]{wayanatavares2005}, containing a reflex of the transitivizer \rc{-ka}, which was restricted to \gl{s_p_} verbs in \PC \parencite{gildea2019overview}.

Further comparative evidence suggests that \rc{ɨpɨtə} \qu{to go down} was originally an \gl{s_p_} verb which switched its class to \gl{s_a_} in some languages, incompletely so in \wayana.
The \arara causativized form is \obj{eniptoŋ} \parencite[66]{alves2017arara}, and \kalina has a cognate form \obj{enɨʔto} \parencite[263]{courtz2008carib}; \obj{onɨʔto} \qu{to go down} in \cref{tab:descend_cog} is a detransitivized form thereof, lit.\ \qu{to get oneself down}.
Both causativized forms contain a reflex of the (rare) transitivizer \rc{en-}, which was usually found with \gl{s_p_} verbs \parencite{gildea2019overview}.
Besides the irregular first person \obj{p-}, \trio \obj{ɨhtə} shows other irregularities, in particular in its causativized forms \parencite[263]{triomeira1999}.
Thus, it seems that this verb was originally \gl{s_p_}, but then switched its class in four and a half languages of the family, for so far unknown reasons.
